---
author: "Angelo Pelonero"
date: "6/3/2019"
title: "Single Cell RNA-Seq Analysis with R Bioconductor" 
output: html_notebook
---

<br/>

###In this workshop, our objectives are to:  

1. Be aware of different methods of loading scRNA-seq data into R
2. Understand the fundamental steps of preprocessing and quality control
3. Identify principal components that identify variability
4. Categorize cell types through clustering
5. Perform non-linear dimension reduction for final evaluation and visualization (e.g. via tSNE and UMAP graphs)

<br/>

#####Load required packages:

Notes: 

* You should already have these packages installed (per setup instructions via email). If you do not, un-comment the `install.packages` line below (but be warned that this may take a while!).   
* If asked "Do you want to install from sources the package which needs compilation?", select Yes [y].  
* Lazy loading errors can be ignored, as can warnings about the version of R packages were built under.  
* If asked to restart R, we recommend that you do so. 

```{r message=FALSE}

## list packages
pkgs <- c("dplyr","cowplot","reticulate","Seurat","umap") 

## install packages
install.packages(pkgs, update = TRUE, ask = FALSE)  

## load packages
for(package in pkgs) {
  do.call("library", list(package))
}
```
To find package citations, you can use the `citation()` function with package name as the argument, e.g. `citation("Seurat")`.  

<br/>

#####Install and load knitr (for R notebook users)

Notes:  

* You should already have this package installed (per setup instructions via email). If you do not, un-comment the `install.packages` line below.  
* If asked to restart R, say No.  
* You can ignore the message "Error in install.packages : Updating loaded packages". This just means you already have it installed/loaded.  

```{r message=FALSE}
install.packages("knitr")
suppressPackageStartupMessages({
  library(knitr)
})
```

For setup questions before 9am on the workshop day, please contact either:    

Angelo Pelonero: angelo.pelonero@ucsf.edu
Karla Lindquist: karla.linquist@ucsf.edu
