---
title: "Single Cell RNA-Seq Analysis - setup" 
authors: "Rebecca Jaszczak, Karla Lindquist, Angelo Pelonero"
date: "3/2/2020"
output: html_notebook
---

<br/>

### In this workshop, our objectives are to:  

1. Be aware of different methods of loading scRNA-seq data into R
2. Understand the fundamental steps of preprocessing and quality control
3. Identify principal components that identify variability
4. Categorize cell types through clustering
5. Perform non-linear dimension reduction for final evaluation and visualization (e.g. via tSNE and UMAP graphs)

<br/>

##### Install and load required packages:

Notes: 

* You should already have these packages installed (per setup instructions via email). If you do not, un-comment the `install.packages` line below (but be warned that this may take a while!).   
* If asked "Do you want to install from sources the package which needs compilation?", select Yes [y].  
* Lazy loading errors can be ignored, as can warnings about the version of R packages were built under. 
* If asked to restart R, we recommend that you do so. 

```{r message=FALSE}

## list packages
pkgs <- c("dplyr","cowplot","Seurat") 

## install packages
install.packages(pkgs, update = TRUE, ask = FALSE)  

## load packages
for(package in pkgs) {
  do.call("library", list(package))
}
```
To find package citations, you can use the `citation()` function with package name as the argument, e.g. `citation("Seurat")`.  

<br/>

##### Install and load knitr (for R notebook users):  

Notes:  

* You should already have this package installed (per setup instructions via email). If you do not, un-comment the `install.packages` line below.  
* If asked to restart R, say No.  
* You can ignore the message "Error in install.packages : Updating loaded packages". This just means you already have it installed/loaded.  

```{r message=FALSE}
#install.packages("knitr")
suppressPackageStartupMessages({
  library(knitr)
})
```

<br/> 

If you have trouble with installations up until the day before the workshop, please email both Angelo and Karla:  
Angelo Pelonero: angelo.pelonero@ucsf.edu  
Karla Lindquist: karla.lindquist@ucsf.edu  

We will also be in the classroom half an hour before the workshop begins. However, this time will be for minor troubleshooting issues only. If you have major issues, or if you wait until this time to begin your installations, we cannot guarantee that you will be set up in time to follow along with the workshop.      

